---
title: Testing Indexes in Postgres & R
author:
  - affiliation: University of Wisconsin - Madison
    name: Simon Goring
date: "September 24, 2019"
output:
  html_document:
    code_folding: show
    fig_caption: yes
    keep_md: yes
    self_contained: yes
    theme: readable
    toc: yes
    toc_float: yes
dev: svg
highlight: tango
---



# Load and Connect the Required Libraries.
```{r}
library(RPostgreSQL)
library(dplyr)
library(purrr)
library(purrrlyr)
library(sf)
library(randomNames)
library(assertthat)
library(sentimentr)

# Build will fail if the file isn't there.
assertthat::assert_that(file.exists('db_connect.txt'),
  msg = "Your connection file is missing.")

con_file <- readr::read_lines('db_connect.txt')

con <- RPostgreSQL::dbConnect(
             PostgreSQL(),
             host = con_file[1],
             port = con_file[2],
             user = con_file[3],
         password = con_file[4],
           dbname = con_file[5])

```

## Generating Fake Data

```{r fakeData}

datatable <- data.frame(oid = 1:1e6,
                        code = NA,
                        value = round(runif(1e6, 10, 100)))

for(i in 1:1e6) {
  # This takes a long time.
  datatable$code[i] <- paste0(letters[sample(1:26, 3, replace = TRUE)], collapse = '')
}

```

```{r testFunction}

test <- function(x) {
  query <- paste0("SELECT * FROM ", x,
                  " WHERE ")
}
dbWriteTable(con,
             "datatablenoindex",
             datatable)

dbWriteTable(con,
             "datatableindex",
             datatable)

```{r}
dbExecute(con,
          "CREATE INDEX testval ON datatable USING HASH code")
```

# How many data pages does this take up?
dbGetQuery(con,
           "SELECT
               pg_relation_filepath(oid),
               relpages
            FROM pg_class
            WHERE relname = 'datatable';")

noindex <- system.time(
  for (i in 1:1000) {
    suppressMessages(dbGetQuery(con, "SELECT *
     FROM datatablenoindex
     WHERE code LIKE 'l%'
        OR code LIKE 'k%';"))
  })

index <- system.time(
  for (i in 1:1000) {
    suppressMessages(dbGetQuery(con, "SELECT *
     FROM datatable
     WHERE code LIKE 'l%'
        OR code LIKE 'k%';"))
  })

ratio <- (index[3] - (sum(index[-3]))) / (noindex[3] - (sum(noindex[-3])))

dbExecute(con,
          "CREATE INDEX testvaldual ON datatable(code, value)")

          noindex <- system.time(
            for (i in 1:1000) {
              suppressMessages(dbGetQuery(con, "SELECT *
               FROM datatablenoindex
               WHERE code LIKE 'l%'
                  AND value > 55;"))
            })

          index <- system.time(
            for (i in 1:1000) {
              suppressMessages(dbGetQuery(con,
                "SELECT *
               FROM datatable
               WHERE code LIKE 'l%'
                  AND value > 55;"))
            })

## Using the HASH

```{r}

dbExecute("CREATE INDEX hashIndex ON datatable USING HASH (value)")

noindexhash <- system.time(
  for (i in 1:1000) {
    suppressMessages(dbGetQuery(con, "SELECT *
     FROM datatablenoindex
     WHERE value = 55;"))
  })

indexhash <- system.time(
  for (i in 1:1000) {
    suppressMessages(dbGetQuery(con,
      "SELECT *
     FROM datatable
     WHERE value = 55;"))
  })
```

## GIN Index

```{r}

dbCreateTable(con,
              "forgin",
              c(oid="serial",
                sentence = "text",
                values = "integer",
                code = "char(3)"))
dbCreateTable(con,
              "forginnix",
              c(oid="serial",
                sentence = "text",
                values = "integer",
                code = "char(3)"))

dbExecute(con,
         "INSERT INTO forgin(sentence, values, code)
         VALUES
         ((SELECT
	substring('ABCDEFGHIJKLMNOPQRSTUVWXYZ'
        FROM (random() *26)::int FOR 1) ||
	substring('abcdefghijklmnopqrstuvwxyz'
        FROM (random() *26)::int FOR 20) AS sentence
      FROM generate_series(1, 10000)),
         (SELECT (random()*1000)::integer FROM generate_series(1, 10000)),
       ())")

```
