---
title: 'Bad Passengers: Where are they at?'
author: "Simon Goring"
date: "September 19, 2019"
output: html_document
---

## Load Libraries

We want to load the libraries

```{r loadLibs}
library(RPostgreSQL)
library(readr)

```

## Make Connection

We're going to onnect to the database.

```{r makeConnect}
connection <- readr::read_lines("db_connect.txt")

con <- RPostgreSQL::dbConnect(PostgreSQL(),
                              host = connection[1],
                              port = connection[2],
                              user = connection[3],
                              password = connection[4],
                              dbname = connection[5])

```

## Question

Which locations have the worst passengers?

* ratings
* rides
* rateobj
* places

## Links

* ratings -> rides using `rideid` (both)
* ratings -> rideobj using `rdid` (both)
* rides -> places using `plid` == `origin` AND `destination`

## Query

```{r ratedPass}

ratePass <- dbGetQuery(con,
                       "SELECT
                          rt.rating,
                          plo.zone AS origin,
                          pld.zone AS destination
                        FROM
                          ratings AS rt
                          INNER JOIN rides AS ri ON rt.rideid = ri.rideid
                          INNER JOIN rateobj AS ro ON ro.rdid = rt.rdid
                          INNER JOIN places AS plo ON plo.plid = ri.origin
                          INNER JOIN places AS pld ON pld.plid = ri.destination
                        WHERE ro.rated LIKE 'Passenger';")
```

Now I have a data table that has `r nrow(ratePass)` rows.  There are `r sum(ratePass$rating == 5)` five star ratings.

```{r}
boxplot(rating ~ origin, data = ratePass)
```

Here is a plot of all the ratings for passengers.

