---
title: Making the Taxi DB
author:
  - affiliation: University of Wisconsin - Madison
    name: Simon Goring
date: "September 16, 2019"
output:
  html_document:
    code_folding: show
    fig_caption: yes
    keep_md: yes
    self_contained: yes
    theme: readable
    toc: yes
    toc_float: yes
dev: svg
highlight: tango
---

## Loading Libraries

We're using a few libraries here:

```{r loadLibs}

library(RPostgreSQL)
library(dplyr)
library(purrr)
library(purrrlyr)
library(sf)
library(randomNames)
library(assertthat)
library(sentimentr)
```

## Loading the database

Create a new local database called `taxico`.  You will connect to it using R (in this document), as I did in the last Rmd document.  First, create a text file called `db_connect.txt` and add the information you need to connect to your local database:

```
host
port
username
password
database
```

We'll use this file to connect to the database:

```{r dbConnect}

# Build will fail if the file isn't there.
assertthat::assert_that(file.exists('db_connect.txt'),
  msg = "Your connection file is missing.")

con_file <- readr::read_lines('db_connect.txt')

con <- RPostgreSQL::dbConnect(
             PostgreSQL(),
             host = con_file[1],
             port = con_file[2],
             user = con_file[3],
         password = con_file[4],
           dbname = con_file[5])

```

## A Taxi Database and Fake Data

Lets say we want to make a database for the taxi company.  We need to create a number of tables, and we need to fill them with fake data.

In class we discussed some of the elements a taxi database might need:

*  Drivers
*  Passengers
*  Dispatchers
*  Locations
*  Ratings (driver, customer)
*  Vehicles
*  Rides
*  Reservations
*  Maintenance
*  Payment [Cash, Debit, Credit, Cheque]

To generate the fake database we're going to use some different datasets.  The first comes from the city of Vancouver:

### Names: Passengers, Dispatchers and Drivers

Random Names: R Library `randomNames` (https://cran.r-project.org/web/packages/randomNames)

We use names in three places (at the most). Data from 2018 indicates that there are approximately 2,100 taxi cabs operating in Vancouver.  Assuming that each cab has two drivers (running at 80% operation/day) then we might expect to have 4,200 total drivers in the Lower Mainland.  Lets assume we're working with one company, with approximately **2,000 drivers**.

I was unable to find reasonable data on passenger information.  I did find a single reference to a total volume of **6,000,000** annual rides for passengers.  Lets assume 200,000 _named_ riders annually.  We can decide how the rides get divvied up later.

Taxi companies employ dispatchers .  Lets make the numbers related and choose **200** as the total number of dispatchers.

```{r fakeNames}

nameTables <- data.frame(people = c("drivers",
                                    "passengers",
                                    "dispatchers"),
                number = c(2000, 200000, 200),
                stringsAsFactors = FALSE)

personName <- nameTables  %>%
  by_row(function(x) {
    randomNames::randomNames(x$number)
  }, .labels = FALSE) %>%
  unlist(recursive = FALSE)

split_name <- function(x) {
  splitName <- strsplit(x, ",")[[1]]

  data.frame(first = splitName[2],
             last  = splitName[1],
             stringsAsFactors = FALSE)
}

# Split apart the names, then add a unique ID.
allNames <- personName %>%
  map(function(x) lapply(x, split_name) %>%
  bind_rows()) %>%
  map(function(x) {
    x$pid <- 1:nrow(x)
    return(x)
  })

names(allNames) <- nameTables$people

drivers <- allNames$drivers %>%
  mutate(province = "BC",
         license = length(pid) %>%
               runif(1e7, 1e8) %>%
               floor,
         renewed = as.Date("2011-01-01") +
                    rgamma(1:2000, 4, 8) *
                    (5 * 365))

drivers$renewed[drivers$renewed >= as.Date(Sys.Date())] <- as.Date(Sys.Date() - 1)

assertthat::assert_that(all(drivers$renewed < as.Date(Sys.Date())))

```

### Places

We're using a map of [Vancouver Neighborhoods (polygons)](https://data.vancouver.ca/datacatalogue/localAreaBoundary.htm).  This uses spatial data in R using the [sf package]().  There are a set of good tutorials for using Spatial Data in R on the [r-spatial Github Pages](https://r-spatial.github.io/sf/articles/sf1.html).

```{r getSpatial}

if (!dir.exists("data/vanhoods/")) {
  if (!dir.exists("data")) {
    dir.create("data")
  }
  dir.create("data/vanhoods")
}

if (!file.exists("data/vanhoods/")) {
  download.file("ftp://webftp.vancouver.ca/OpenData/shape/local_area_boundary_shp.zip",
  "data/vanhoods/local_area_boundary_shp.zip")
  unzip("data/vanhoods/local_area_boundary_shp.zip",
        exdir = "data/vanhoods")
}

places <- sf::st_read("data/vanhoods/local_area_boundary.shp")

plot(places)
```

We said we wanted something like 6,000,000 annual rides.  That means 6M ride starts and 6M ride ends.  The `sf` package has a function called [`st_sample()`](https://r-spatial.github.io/sf/reference/st_sample.html) that lets us generate random samples within polygons:

```{r addLocs}
# EPSG: 26910
# +proj=utm +zone=10 +datum=NAD83 +units=m +no_defs
n_hoods <- nrow(places[1])

locs <- data.frame(coords = st_as_text(sf::st_sample(places,
                            rep(20000,n_hoods))),
                   zone = rep(as.character(places$NAME),
                       each = 20000),
                       stringsAsFactors = FALSE)
locs$plid <- 1:nrow(locs)

```

### Maintenance

In British Columbia all vehicles must be inspected twice yearly.  We need to generate fake car information.  A car is made up of information about the model, the year and the license plate.

```{r genCars}

no_cars <- 2100

car_make <- data.frame(modelid = 1:4,
                       make = c("Ford", "Fiat",
                                "SAIC", "Tata"),
                       model = c("Fiesta", "500",
                                 "Roewe", "Tigor"),
                       stringsAsFactors = FALSE)

lsa <- function(x) { sample(letters, x,
                            replace = TRUE) }

nsa <- function(x) { sample(0:9, x, replace = TRUE) }

cars <- data.frame(carid = 1:no_cars,
                   modelid = sample(car_make$modelid,
                                    2100,
                                  replace = TRUE),
                   license = paste0(lsa(2100),
                                    nsa(2100),
                                    lsa(2100),
                                    nsa(2100),
                                    lsa(2100),
                                    nsa(2100)))
```

### Rides

Rides link two locations, a passenger, a car and a driver.

```{r buildRide}

date_seq <- seq(from = as.Date("2014-01-01"),
                  to = as.Date("2019-01-01"),
                  by = 1)

date_prob <- (date_seq %>%
              lubridate::wday()) %in% c(1,7) %>%
              if_else(., 0.3, 0.7)

ride_date <- sort(sample(date_seq,
                         size = 20000,
                         replace = TRUE,
                       prob = date_prob))


rideTable <- 1:20000 %>%
  map(function(x) {
    spots <- sample(nrow(locs), 2, replace = FALSE)
    data.frame(origin = spots[1],
              destination = spots[2],
              driver = sample(drivers$pid, 1),
              rider = sample(allNames$passengers$pid, 1))
  }) %>%
  bind_rows() %>%
  mutate(date = ride_date,
         rideid = 1:20000)

```

### Ride Ratings

Lets assume that we can have either passenger or driver ratings.  We'll make a table that has two options, and a object ID.  Then, we will randomly assign ratings for each ride.  We will first decide if people leave a rating (assume a 60% chance someone will rate their driver, and an 80% chance a driver rates a passenger).

```{r ratings}

sents <- sentimentr::combine_data("hotel_reviews") %>% select(text) %>% lapply(function(x)strsplit(x, ".")) %>% unlist()

sents <- sents[nchar(sents) > 10]

rateobj <- data.frame(rdid = c(1,2),
                      rated = c("Passenger", "Driver"))

ratings <- data.frame(rideid = rep(1:20000, 2),
                      rdid  = rep(1:2, each = 20000),
                      assigned = runif(40000, 0, 1),
                      rating = sample(5,40000, replace = TRUE, prob = ppois(1:5, 3)),
                    comments = NA)

  ratings$comments[ratings$assigned > .9] <- sents[sample(1:length(sents),
               sum(ratings$assigned > .9),
             replace = TRUE)]

ratings <- ratings %>%
  filter((rdid == 1 & assigned < .8) | (rdid == 2 & assigned < .6)) %>%
  select(-assigned)

```

## Pushing things to the DB

We need to add the tables now. For each table we're going to check whether or not the table exists.  If it exists we're going to write the table to file.

```{r writeTables}
# Note: With all this repetition we should make a function.

if (!dbExistsTable(con, "drivers")) {
  dbWriteTable(con, "drivers", drivers,
row.names = FALSE)
}

if (!dbExistsTable(con, "passengers")) {
  dbWriteTable(con, "passengers",
               allNames$passengers,
             row.names = FALSE)
}

if (!dbExistsTable(con, "dispatchers")) {
  dbWriteTable(con,
               "dispatchers",
               allNames$dispatchers,
             row.names = FALSE)
}

if(!dbExistsTable(con, "places")) {
  aa <- try(dbExecute(con, "CREATE EXTENSION postgis;"))
  dbWriteTable(con,
               "places",
               locs,
             row.names = FALSE)
}

if (!dbExistsTable(con, "rides")) {
  dbWriteTable(con, "rides", rideTable,
row.names = FALSE)
}

if (!dbExistsTable(con, "model")) {
  dbWriteTable(con, "model", car_make,
row.names = FALSE)
}

if (!dbExistsTable(con, "cars")) {
  dbWriteTable(con, "cars", cars,
row.names = FALSE)
}

if (!dbExistsTable(con, "ratings")) {
  dbWriteTable(con, "ratings", ratings,
row.names = FALSE)
}

if (!dbExistsTable(con, "rateobj")) {
  dbWriteTable(con, "rateobj", rateobj,
row.names = FALSE)
}

dbDisconnect(con)
```

## There are no Explicit Relationships Here!

We're going to add them in and then time the differences.  We can also test things with the ratings, and potentially with distances.

## Questions

### Which pickup/dropoff locations have the worst customers?
