---
title: "MBAN_507 Connecting to Datbases"
author: "Your Name Here"
date: "September 5, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Connecting to Databases

To connect to a database you need to know the location of the database (IP Address), you need to know the server port through which the database can be accessed (most postgres databases use **5432**), you also need to know the user name and password associated with your account on the database.

To manage and access data in R we have to use a package designed to return data from the database into data formats that can be recognized by R.  So the package communicates between R and the database.  Anything that does this is said to use an Application Program Interface, or API.

In R, we can use `RPostgreSQL`, a package designed specifically for interfacing with Postgres databases.

```{r loadLibraries}
library(RPostgreSQL)

```

You will see that the package also loads another package called `DBI`.

## Connecting to the Database

The library gives us the tools to interact with the database, but we need to tell R where the database connection actually is.  We need to know the:

Element | Value
------ | -----
Host | mban-bait.cmy5mce9ahmj.us-east-1.rds.amazonaws.com
Port | 5432
User | **supersecret**
Password | **supersecret**
Database | mbandatabase


```{r connecting, eval=FALSE}
con <- dbConnect(PostgreSQL(),
                 host = "mban-bait.cmy5mce9ahmj.us-east-1.rds.amazonaws.com",
                 port = 9999999,
                 user = "XXXXXXXXXXXX",
                 password = "XXXXXXXXXXXXXX",
                 dbname = "XXXXXXXXXXXXXXXXXX")
```

If we did this properly we now have an object `con` in R that represents the connection to our database.  From then on, any time we want to explicitly connect to the database we can use the object `con`.

For example, if we wanted to list all the tables, we could use the command:

```{r listTables, eval=FALSE}
dbListTables(con)
```

There are a number of basic commands you can do to manage a database connection with the `RPostgreSQL` package.

Function | Description
------ | -------
`dbGetInfo()` | Get information (as a `list`) about the current connection.
`dbReadTable()` | Read information from a table in the database.
`dbSendQuery()` | Send a SQL query to the database.
`dbFetch()` | Fetch the results from the database.

## Our First Data Manipulation

There is one table in the database.  Use this chunk to read the table by replacing the `data.frame()` call with the appropriate call.

```{r readTable}

table_data <- data.frame(rows = c(1,1))

```

When we do this we see that the table has `r nrow(table_data)` unique entries.

We can get subsets of data by filtering out categories.  For example, I am only interested in data related to "University professors and lecturers":

```{r filtering, eval = FALSE}
library(dplyr)

profs <- table_data %>%
  filter(Description == "University professors and lecturers")


```

This now gives us `r 12` rows out of our original `data.frame()`.

The data in the original CSV was not well formatted, and so we have to mess around a bit to get it to plot properly.  But we can do it:

```{r plotprofs, eval = FALSE}
plot(seq(2018, 2028, by = 1), as.numeric(profs[1,6:16]),
     type = "b",
     xlab = "Year",
     ylab = "Employment")

```

## Summary

Please use this space to write some summary statements.
